### 2016-05-31 Paper_v1.tex
May31

\subsection*{Estimation of an ensmeble of models}
We constructed an initial model of cell-free \textit{E. coli} metabolism by removing the growth-associated processes from the genome-scale model of Palsson and coworkers \cite{?} and adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset from Swartz and coworkers \cite{?}. 
This initial model captured the core metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids, and CAT.
We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term.
We used multiple saturation kinetics to model metabolic fluxes and mass action kinetics to model enzyme degradation.
We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function \cite{?}, and the control term was calculated as the mean of all transfer functions.
%The model was trained against the Swartz dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
The model was trained against the Swartz dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (AXP, GXP, CXP, UXP), and 18 of the 20 proteinogenic amino acids.
We generated an ensemble of 18,000 parameter sets by minimizing the error between the Swartz dataset and the metabolite concentrations predicted by the model.
Out of these 18,000 sets, we defined the set with the lowest error value as our best-fit set.
The ensemble captures the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates (Fig.~\ref{fig:BothCarbon}, top).
Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:BothCarbon}, bottom).
The ensemble also captures the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base (Fig.~\ref{fig:Energy}).
The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it. Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.

\subsection*{Sensitivity analysis}
We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
This is explained by GTP and the amino acids being reactants for CAT synthesis.
Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
This is because an increase in saturation constant causes a decrease in the corresponding reaction rate \cite{?}.
The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is not the limiting factor to CAT production.

@@

\subsection*{Calculation of CAT yield}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
Arginine and glutamate were excluded due to not being present in the Swartz dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield of 0.1942.
Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 45\% of the theoretical maximum.
This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method \cite{?}.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes a weight, equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere, $\mathcal{T}$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 652, which includes 163 rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, set to 0.03, and $r$ denotes a random sample from the normal distribution.
We stored the parameter set and calculated its cost; if it was less than the previous cost, we used the new parameter set to generate the following set.
After generating 180,000 sets we defined the 18,000 sets with the lowest cost values as our ensemble, and the set with the lowest cost value as our best-fit set.

>>>>

\subsection*{Estimation of an ensmeble of models}
We constructed an initial model of cell-free \textit{E. coli} metabolism by removing the growth-associated processes from the genome-scale model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS} and adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}. 
This initial model captured the core metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids, and CAT.
We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term.
We used multiple saturation kinetics to model metabolic fluxes and mass action kinetics to model enzyme degradation.
We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function, and the control term was calculated as the mean of all transfer functions.
%The model was trained against the training dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
The model was trained against the training dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (AXP, GXP, CXP, UXP), and 18 of the 20 proteinogenic amino acids.
We generated an ensemble of 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model.
Out of these 18,000 sets, we defined the set with the lowest error value as our best-fit set.
The ensemble captures the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates (Fig.~\ref{fig:BothCarbon}, top).
Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:BothCarbon}, bottom).
The ensemble also captures the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base (Fig.~\ref{fig:Energy}).
The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it. Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.

\subsection*{Sensitivity analysis}
We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
This is explained by GTP and the amino acids being reactants for CAT synthesis.
Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is not the limiting factor to CAT production.

@@

\subsection*{Calculation of CAT yield}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield of 0.1942.
Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 45\% of the theoretical maximum.
This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes a weight, equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere, $\mathcal{T}$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 652, which includes 163 rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, set to 0.03, and $r$ denotes a random sample from the normal distribution.
We stored the parameter set and calculated its cost; if it was less than the previous cost, we used the new parameter set to generate the following set.
After generating 180,000 sets we defined the 18,000 sets with the lowest cost values as our ensemble, and the set with the lowest cost value as our best-fit set.
